---
title: "Descriptive analysis of statistical associations with descriptio"
author: "Nicolas Robette"
date: "`r Sys.Date()`"
output: 
  rmdformats::html_clean:
    thumbnails: FALSE
    use_bookdown: FALSE
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

oldpar <- par() 
oldoptions <- options()

## Global options
options(max.print="75", width=1000)
opts_chunk$set(collapse = TRUE,
		           comment = "#>",
		           echo=TRUE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=FALSE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

\ 

`descriptio` package provides some functions dedicated to the description of statistical associations between variables. They are based on effect size measures (also called association measures).

The main measures are built from simple concepts (correlations, proportion of variance explained), they are bounded (between -1 and 1 or between 0 and 1) and are not sensitive to the number of observations.

The main measures of *global* association are the following.

- For the relationship between two categorical variables: the **Cramér's V** which, unlike the chi-square, for example, is not sensitive to the number of observations or the number of categories of the variables. It varies between 0 (no association) and 1 (perfect association). Squared, it can be interpreted as the share of variation shared between two variables.

- For the relationship between two numerical variables: Kendall's (*tau*) or Spearman's (*rho*) **rank correlations**, which detect *monotonic* relationships between variables, and not only linear ones as is the case with Pearson's linear correlation. They vary between -1 and 1. An absolute value of 0 indicates no association, an absolute value of 1 a perfect association. The sign indicates the direction of the relationship.

- For the relationship between a categorical variable and a numerical variable: the square of the **correlation ratio** (*eta²*). It expresses the proportion of the variance of the numerical variable "explained" by the categorical variable and varies between 0 and 1.

In addition to measures of global association, we also use measures of *local* association, i.e. at the level of the categories of the variables.

- For the relationship between two categorical variables: the **phi coefficient** measures the attraction or repulsion in a cell of a contingency table. It varies between -1 and 1. An absolute value of 0 indicates an absence of association, an absolute value of 1 a perfect association. There is attraction if the sign is positive, repulsion if the sign is negative. Squared, phi is interpreted as the proportion of variance shared by the two binary variables associated with the categories studied. Unlike the *test value*, phi is not sensitive to the sample size.

- For the relationship between a categorical variable and a numerical variable: the **point biserial correlation** measures the magnitude of the difference between the means of the numerical variable according to whether or not one belongs to the category studied. It varies between -1 and 1. An absolute value of 0 indicates no association, an absolute value of 1 a perfect association. The sign indicates the direction of the relationship. When squared, point biserial correlation can be interpreted as the proportion of variance of the numerical variable "explained" by the category of the categorical variable.

Note that if we code the categories of the categorical variables as binary variables with values of 0 or 1, the phi coefficient and the point biserial correlation are equivalent to Pearson's correlation coefficient.

For more details on these effect size measurements, see:
Rakotomalala R., [« Comprendre la taille d'effet (effect size) »](http://eric.univ-lyon2.fr/~ricco/cours/slides/effect_size.pdf)  
\ 

In some functions of `descriptio`, association measures can be completed by **permutation tests**, which are part of *combinatorial inference* and constitute a *nonparametric* alternative to the significance tests of frequentist inference. A permutation test is performed in several steps.

1. A measure of association between the two variables under study is computed.

2. The same measure of association is calculated from a "permuted" version of the data, i.e. by randomly "mixing" the values of one of the variables, in order to "break" the relationship between the variables.

3. Repeat step 2 a large number of times. This gives an *empirical* distribution (as opposed to the use of a *theoretical* distribution by frequentist inference) of the measure of association under the H0 hypothesis of no relationship between the two variables.

4. The result of step 1 is compared with the distribution obtained in 3. The *p-value* of the permutation test is the proportion of values of the H0 distribution that are more extreme than the measure of association observed in 1.

If all possible permutations are performed, the permutation test is called "exact". In practice, the computation time required is often too important and only a part of the possible permutations is performed, resulting in an "approximate" test. In the following examples, the number of permutations is set to 100 to reduce the computation time, but it is advisable to increase this number to obtain more accurate and reliable results (for example `nperm`=1000).  
\ 

To illustrate the statistical association analysis functions of `descriptio`, we will use data on cinema. This is a sample of 1000 films released in France in the 2000s, for which we know the budget, the genre, the country of origin, the "art et essai" label, the selection in a festival (Cannes, Berlin or Venice), the average rating of intellectual critics (according to Allociné) and the number of admissions. Some of these variables are numerical, others are categorical.

```{r load}
library(descriptio)
data(Movies)
str(Movies)
```


----

# Relationship between two variables

The package offers several functions to study the statistical relationship between two variables, depending on the nature (categorical or numerical) of these variables.

### Two categorical variables

The function `assoc_twocat` computes :

- the contingency table (numbers)
- the percentages, the row-percentages and the column-percentages
- the theoretical numbers, i.e. in a situation of independence
- the chi-square 
- the Cramér's V and the p-value of the corresponding permutation test
- the Goodman & Kruskal tau (an asymetrical measure of association)
- the Pearson residuals, adjusted or not
- the odds ratios for every cell of the contingency table
- the phi coefficients and the p-values of the corresponding permutation tests
- the global and local PEMs (Percentage of Maximum Deviation from Independence, see Cibois 1993)
- a summary table of these results

```{r assoc_twocat}
res <- assoc.twocat(Movies$Country, Movies$ArtHouse, nperm=100)
res$tables
res$global
res$local
res$gather
```

\
The function `ggassoc_crosstab` presents the contingency table in graphical form, with rectangles whose area corresponds to the number of observations and whose color gradient corresponds to the attraction/repulsion (from one of the local association measures proposed in `assoc.twocat`, here the phi coefficients). The "art et essai" label is clearly over-represented among French films and under-represented among American films.

```{r ggassoc_crosstab}
ggassoc_crosstab(Movies, ggplot2::aes(x=Country, y=ArtHouse))
```

The function `ggassoc_phiplot` proposes another way to represent the attractions/repulsions. The width of the rectangles corresponds to the numbers of the variable x, their height to the local associations (by default phi coefficients). The rectangles are colored in black when there is attraction, in white when there is repulsion.

```{r ggassoc_phiplot}
ggassoc_phiplot(Movies, ggplot2::aes(x=Country, y=ArtHouse))
```

The function `ggassoc_bertin` is inspired by Jacques Bertin's principles of graphical semiology for the representation of a data matrix and by the tool [AMADO Online](https://paris-timemachine.huma-num.fr/amado/main.html). The height of the bars corresponds to the conditional probabilities of y given x. They are colored black when the conditional probabilities are greater than the marginal probabilities of y. The width of the bars can be adjusted to be proportional to the marginal probabilities of x.

```{r ggassoc_bertin}
ggassoc_bertin(Movies, ggplot2::aes(x=Country, y=ArtHouse), prop.width = TRUE, add.gray = TRUE)
```

The function `ggassoc_assocplot` produces an "association plot" as proposed by Cohen (1980) and Friendly (1992). The width of the bars is proportional to the square root of the theoretical numbers. The height of the bars and the color gradient are proportional to the local associations. If these local associations are measured by Pearson residuals (the default choice), the area of the bars is proportional to the difference between the theoretical and observed numbers. 

When the number of categories of the variables is high, as here when we cross the genre of the film and its geographical origin, it can be useful to sort the rows and/or the columns so that those which are similar are close.

We see that action, animation, science fiction and horror films are over-represented among American films, and that documentaries, dramas and comedy-dramas are over-represented among French films.

```{r ggassoc_assocplot}
ggassoc_assocplot(Movies, ggplot2::aes(x=Country, y=Genre), sort = "both")
```

The function `ggassoc_chiasmogram` draws a "chiasmogram", a graphical representation proposed by Bozon and Héran (1988). The width of the rectangles is proportional to the marginal probabilities of the variable in column, their height is proportional to the marginal probabilities of the variable in row. The surface area of the rectangles is therefore proportional to the theoretical numbers. The rectangles are colored according to the degree of local association (by default the phi coefficients).

```{r ggassoc_chiasmogram}
ggassoc_chiasmogram(Movies, ggplot2::aes(x=Country, y=Genre), sort = "both")
```

The function `ggassoc_marimekko` produces marimekko charts, also called mosaic plots. The width of the bars is proportional to the marginal probabilities of the variable x, their height to the conditional probabilities of the variable y given x.

```{r ggassoc_marimekko1}
ggassoc_marimekko(Movies, ggplot2::aes(x=Country, y=Genre), sort = "both", type = "classic")
```

We can choose to color the bars according to the degree of local association (here the phi coefficients), as proposed by Friendly (1994).

```{r ggassoc_marimekko2}
ggassoc_marimekko(Movies, ggplot2::aes(x=Country, y=Genre), sort = "both", type = "shades")
```

We can also choose to replace the gradient of colors by textures more or less dense according to the degree of local association.

```{r ggassoc_marimekko3}
ggassoc_marimekko(Movies, ggplot2::aes(x=Country, y=Genre), sort = "both", type = "patterns")
```



### One categorical variable and one numerical variable

The `assoc_catcont` function computes:

- the squared correlation ratio (eta²) and the p-value of the corresponding permutation test
- the point biserial correlations and the p-values of the corresponding permutation tests

```{r assoc_catcont}
assoc.catcont(Movies$Country, Movies$Critics, nperm=100)
```

\
The function `ggassoc_boxplot` represents the relationship between the variables in the form of box-plots and/or "violin" distributions.

```{r ggassoc_boxplot, eval = FALSE}
ggassoc_boxplot(Movies, ggplot2::aes(x=Country, y=Critics))
```



### Two numerical variables

The function `assoc_twocont` computes the Kendall and Spearman rank correlations and the Pearson linear correlation, as well as the p-values of the corresponding permutation tests.

```{r assoc_twocont}
assoc.twocont(Movies$Budget, Movies$BoxOffice, nperm=100)
```

\
The function `ggassoc_scatter` represents the relationship between the two variables in the form of a scatterplot, with an approximation by smoothing (with the "Generalized Additive Model" method).

```{r ggassoc_scatter, eval = FALSE}
ggassoc_scatter(Movies, ggplot2::aes(x=Budget, y=BoxOffice))
```



# Relationships between one variable Y and a set of variables X

Often, not just two variables are studied, but a larger set of variables. When one of these variables has the status of an "explained" variable, one generally uses regression models or, possibly, supervised learning models (see the vignette of the `moreparty` package for an example). However, it is essential to know all the bivariate relations of the dataset before moving to an "all else being equal" approach. 

It should be noted that if we do this work in a meticulous way, adding eventually the descriptive analysis of the relationships between three or four variables, we often realize that the surplus of knowledge brought by the regression models is quite limited.  
\

The function `assoc.yx` computes the global association between Y and each of the variables of X, and for all pairs of variables of X.

```{r assoc_yx}
assoc.yx(Movies$BoxOffice, Movies[,-7], nperm=10)
```

\
The functions `catdesc` and `condesc` allow to go into more detail about the relationships, by going to the category level.

`catdesc` deals with the cases where Y is a categorical variable. For a categorical variable X1, it computes, for a given category of Y and a category of X1 :

- the percentage of the category of Y in the category of X1
- the percentage of the X1 category in the Y category
- the percentage of the X1 category in the whole sample
- the local association (by default the phi coefficient)

The results are sorted by decreasing local association and can be filtered to keep only the associations above a given threshold (in absolute value).

For a numerical variable X2, for a given category of Y, it computes :

- the median of X2 in the category of Y
- the median of X2 in the whole sample
- the dispersion of X2 in the Y category
- the dispersion of X2 in the whole sample
- the point biserial correlation

Dispersion is measured by the "median absolute deviation" (MAD), which is the median of the absolute deviations from the median. The median and MAD are so-called "robust" indicators, i.e. not sensitive to outliers, but one can choose to use the mean and standard deviation instead.

```{r catdesc}
res <- catdesc(Movies$Festival, Movies[,-5], nperm = 10)
res$bylevel$Yes
```

\
`condesc` deals with the cases where Y is a numerical variable. For a categorical variable X1, it computes :

- the median of Y in the category of X1
- the median of Y in the whole sample
- the dispersion of Y in the category of X1
- the dispersion of Y in the whole sample
- the point biserial correlation

The results are sorted by decreasing point biserial correlation and can be filtered to keep only associations above a given threshold (in absolute value).

For the numerical variables of X, it calculates the correlation (by default Kendall's tau).

```{r condesc}
condesc(Movies$BoxOffice, Movies[,-7], nperm=100)
```

\
The `darma` function presents the results in a form close to that of a table of results of a regression.

When the variable Y is numerical, the function computes :

- for numerical X variables, the correlation (by deafult Kendall's tau) and the p-value of the corresponding permutation test
- for the categories of the categorical X variables, the median of Y in the category, the point biserial correlation and the p-value of the corresponding permutation test

```{r darma1, eval=FALSE}
res <- darma(Movies$BoxOffice, Movies[,-7], nperm=10)
```

```{r darma1b, echo=FALSE}
knitr::kable(res, row.names=FALSE, format="simple")
```

\
When the variable Y is categorical, the function computes :

- for numerical X variables, the point biserial correlation and the p-value of the corresponding permutation test
- for the categories of the categorical X variables, the percentage of Y in the category, the phi coefficient and the p-value of the corresponding permutation test.

```{r darma2, eval=FALSE}
res <- darma(Movies$Festival, Movies[,-5], target=2, nperm=10)
```

```{r darma2b, echo=FALSE}
knitr::kable(res, row.names=FALSE, format="simple")
```


# Relationships between all variables in a set

Finally, the `ggassoc_*` functions are designed to be integrated in the plot matrices of the [`GGally`](https://ggobi.github.io/ggally/) package. It is thus possible to use them to represent in a single plot all the bivariate relations of a group of variables. 

```{r ggally}
library(GGally)
ggpairs(Movies,
        lower = list(continuous = ggassoc_scatter,
                     combo = ggassoc_boxplot,
                     discrete = ggassoc_crosstab),
        upper = list(continuous = ggassoc_scatter,
                     combo = ggassoc_boxplot,
                     discrete = ggassoc_crosstab),
        diag = list(continuous = wrap("diagAxis", gridLabelSize = 3),
                           discrete = wrap("diagAxis", gridLabelSize = 3)))
```

